---
title: "PhenotypeR"
subtitle: "Review codelists and cohorts in OMOP CDM"
format:
  revealjs: 
    theme: [simple, ../../styleSS25.scss]
    incremental: true   
    slide-number: true
    chalkboard: true
    preview-links: auto
    margin: 0.07
    code-link: true
    code-line-numbers: false
    height: 900
    width: 1600
    footer: |
      <div style="position: relative; width: 100%;">
        <div style="text-align: center; font-weight: 500;">Real World Evidence Summer School 2025</div>
        <div style="position: absolute; right: 20px; top: -800px;"><a href="../index.qmd" style="text-decoration: none;">⬅ Back to Practicals</a></div>
      </div>
execute:
  echo: true
  eval: true
editor: visual
---

## Introduction

-   PhenotypeR package is designed to review cohorts in OMOP CDM.

::: {style="margin-bottom: 10px;"}
:::

-   The code is publicly available in OHDSI's GitHub repository [PhenoypeR](https://github.com/OHDSI/PhenotypeR).

::: {style="margin-bottom: 10px;"}
:::

-   PhenotypeR 0.1.5 is available in [CRAN](https://cran.r-project.org/web/packages/PhenotypeR/index.html){.link}.

::: {style="margin-bottom: 10px;"}
:::

-   Vignettes with further information can be found in the package [website](https://ohdsi.github.io/PhenotypeR/){.link}.

## Set of Functions: Individual Diagnostics Assessment

::: {style="margin-bottom: 10px;"}
:::

::::::: columns
:::: {.column width="50%"}
-   **Database diagnostics**

    -   Summarises the databases metadata and the observation period.
    -   `databaseDiagnostics()`

::: {style="margin-bottom: 10px;"}
:::

-   **Codelist diagnostics**

    -   Summarises the codelist use associated to our cohort.
    -   `codelistDiagnostics()`
::::

:::: {.column width="50%"}
-   **Cohort diagnostics**

    -   Summarises baseline characteristics, attrition, overlap, and large-scale characteristics. It also allows to compare our results with a matched cohort.
    -   `cohortDiagnostics()`

::: {style="margin-bottom: 10px;"}
:::

-   **Population diagnostics**

    -   Incidence and prevalence of our cohort in the database.
    -   `populationDiagnostics()`
::::
:::::::

## Set of Functions: Phenotype Diagnostics
-   Comprises all the diagnostics that are being offered in this package.

. . .

```{r, eval = FALSE}
result <- phenotypeDiagnostics(
  cohort,
  databaseDiagnostics = TRUE,
  codelistDiagnostics = TRUE,
  cohortDiagnostics = TRUE,
  match = TRUE,
  matchedSample = NULL,
  populationDiagnostics = TRUE,
  populationSample = NULL,
  populationDateRange = as.Date(c(NA,NA))
)
```

-   Create a shiny app to visualize all the results

. . .

```{r, eval = FALSE}
shinyDiagnostics(result, directory = here())
```

# Run Phenotype Diagnostics - Example
```{r, eval = FALSE}
install.packages("PhenotypeR")

library(CDMConnector)
library(CodelistGenerator)
library(CohortConstructor)
library(CohortCharacteristics)
library(dplyr)
library(PhenotypeR)
library(here)

requireEunomia(datasetName = "synpuf-1k", cdmVersion = "5.3")

con <- DBI::dbConnect(duckdb::duckdb(), dbdir = eunomiaDir(datasetName = "synpuf-1k", cdmVersion = "5.3"))
cdm <- cdmFromCon(
  con, 
  cdmSchema = "main", 
  writeSchema = "main",
  writePrefix = "my_study_",
  achillesSchema = "main"
)

drug_codes <- getDrugIngredientCodes(
  cdm = cdm, 
  name = c("diclofenac", "acetaminophen"),
  nameStyle = "{concept_name}"
)

cdm$medications <- conceptCohort(
  cdm = cdm, 
  conceptSet = drug_codes, 
  name = "medications"
)

# Run Phenotype Diagnostics 
result <- phenotypeDiagnostics(cdm$medications,
                     databaseDiagnostics = TRUE,
                     codelistDiagnostics = TRUE,
                     cohortDiagnostics = TRUE,
                     matchedAnalysis = TRUE,
                     matchedSample = NULL,
                     populationDiagnostics = TRUE,
                     populationSample = NULL,
                     populationDateRange = as.Date(c(NA,NA)))

shinyDiagnostics(result, here::here())
```

# Visualise the results 
See the results in the [shiny app](https://dpa-pde-oxford.shinyapps.io/PhenotypeR_RWESummerSchool/)
