---
title: "CohortSurvival"
subtitle: "Conduct survival analyses"
format:
  revealjs: 
    theme: [simple, ../../styleSS25.scss]
    incremental: true   
    slide-number: true
    chalkboard: true
    preview-links: auto
    margin: 0.07
    code-link: true
    code-line-numbers: false
    height: 900
    width: 1600
    footer: |
      <div style="position: relative; width: 100%;">
        <div style="text-align: center; font-weight: 500;">Real World Evidence Summer School 2025</div>
        <div style="position: absolute; right: 20px; top: -800px;"><a href="../index.qmd" style="text-decoration: none;">⬅ Back to Practicals</a></div>
      </div>
execute:
  echo: true
  eval: true
editor: visual
---

## Introduction

-   CohortSurvival is a package designed to support descriptive survival studies in R, using data mapped to the OMOP Common Data Model.

::: {style="margin-bottom: 10px;"}
:::

-   The code is publicly available in the DARWIN EU's GitHub repository [CohortSurvival](https://github.com/darwin-eu/CohortSurvival).

::: {style="margin-bottom: 10px;"}
:::

-   CohortSurvival v1.0.1 is available in [CRAN](https://cran.r-project.org/web/packages/CohortSurvival/CohortSurvival.pdf){.link}.

::: {style="margin-bottom: 10px;"}
:::

-   Vignettes with further information can be found in the package [website](https://ohdsi.github.io/CohortSurvival/index.html){.link}.

## How do you use CohortSurvival?

. . .

::: {style="margin-bottom: 15px;"}
:::

**1) Create a cdm object with the cohorts of interest**

You only need to specify the exposure and the outcome cohort names.

. . .

::: {style="margin-bottom: 25px;"}
:::

**2) Run the survival estimation**

Add any additional strata or input parameters necessary for your study.

. . .

::: {style="margin-bottom: 25px;"}
:::

**3) Plot and tabulate the results**

Use the in-built visualisation functions from the package to plot the survival estimates and display the survival summary in a neat table.

## CohortSurvival's main functionality

:::::::::::::: columns
::: {.column width="10%"}
:::

:::::: {.column width="40%"}
::::: bulletbox
:::: {.fragment .fade-in-then-semi-out}
::: {layout="[[-1], [1], [-1]]"}
**Estimate single event survival** *Specify exposure and outcome cohorts.*
:::
::::
:::::
::::::

::: {.column width="5%"}
:::

:::::: {.column width="40%"}
::::: bulletbox
:::: {.fragment .fade-in-then-semi-out}
::: {layout="[[-1], [1], [-1]]"}
**Estimate competing risk survival** *Specify exposure, outcome and competing outcome cohorts.*
:::
::::
:::::
::::::

::: {.column width="5%"}
:::
::::::::::::::

:::::::: columns
::: {.column width="30%"}
:::

:::::: {.column width="40%"}
::::: bulletbox
:::: {.fragment .fade-in-then-semi-out}
::: {layout="[[-1], [1], [-1]]"}
**Do further survival analyses** *Add survival information to the cohort of interest to run survival models using other well-known packages (i.e. survival).*
:::
::::
:::::
::::::
::::::::

# Estimate survival

## Create cdm and get cohorts

```{r, echo=FALSE}
# Load relevant packages (make sure you have the latest versions of all of them)
library(CDMConnector)
library(DrugUtilisation)
library(dplyr)
library(gt)
library(CohortSurvival)
library(ggplot2)
```

```{r}
# Download Eunomia 
requireEunomia()

# Connect to the "database"
con <- DBI::dbConnect(duckdb::duckdb(), dbdir = eunomiaDir())
# Create CDM reference object
cdm <- cdmFromCon(
  con, 
  cdmSchema = "main", 
  writeSchema = "main",
  writePrefix = "my_study_"
)

```

## Use estimateSingleSurvival() for treatment adherence

Single event survival estimation function, with all its input parameters.

```{r, eval = FALSE}
estimateSingleEventSurvival(
  cdm,
  targetCohortTable,
  outcomeCohortTable,
  outcomeDateVariable = "cohort_start_date",
  outcomeWashout = Inf,
  censorOnCohortExit = FALSE,
  censorOnDate = NULL,
  followUpDays = Inf,
  strata = NULL,
  eventGap = 30,
  estimateGap = 1,
  restrictedMeanFollowUp = NULL,
  minimumSurvivalDays = 1
)
```

## Use estimateSingleSurvival() for treatment adherence

Plotting function, with all its input parameters.

```{r, eval = FALSE}
plotSurvival(
  result,
  ribbon = TRUE,
  facet = NULL,
  colour = NULL,
  cumulativeFailure = FALE,
  riskTable = FALSE,
  riskTable(30)
)
```

## Use estimateSingleSurvival() for treatment adherence

Main tabulating function, with all its input parameters.

```{r, eval = FALSE}
tableSurvival(
  x,
  times = NULL,
  timeScale = "days",
  header = c("estimate"),
  type = "gt",
  groupColumn = NULL,
  .options = list()
)
```

## Use estimateSingleSurvival() for treatment adherence

In our case, we will study discontinuation of ibuprofen. Therefore, we will use the same cohort as both target and outcome. Additionally, we will need to change `outcomeDateVariable` to `cohort_end_date`.

. . .

```{r}
cdm <- generateIngredientCohortSet(
  cdm = cdm,
  name = "ibuprofen",
  ingredient = "ibuprofen",
  gapEra = 7
)

survivalResult <- estimateSingleEventSurvival(
  cdm = cdm, 
  targetCohortTable = "ibuprofen",
  outcomeCohortTable = "ibuprofen",
  outcomeDateVariable = "cohort_end_date"
)
```

## Use estimateSingleSurvival() for treatment adherence

```{r}
glimpse(survivalResult, width = 100)
```

## Use estimateSingleSurvival() for treatment adherence

```{r}
plotSurvival(survivalResult)
```

. . .

```{r}
tableSurvival(survivalResult, times = c(7, 30, 75))
```

## Use estimateSingleSurvival() for treatment adherence

::: {style="margin-bottom: 20px;"}
:::

You can also visualise the risk table, with information on number of people at risk, number of events and number of people censored by event timepoint (defined by `eventGap`).

```{r}
riskTable(survivalResult)
```

# Your turn!

## Exercise 1 - Estimate treatment adherence of aspirin users

**Create a cohort of aspirin use and estimate treatment adherence.**

-   How many people are at risk at cohort entry, and how many are left at risk after a month?
-   Plot the drug discontinuation curve and compare it to the ibuprofen one. If you can, plot both Kaplan-Meier curves in the same survival plot.

. . .

*Hint:* use `omopgenerics::bind()` and the `colour` option in `plotSurvival()` for the two outcomes.

## Exercise 1 - Estimate treatment adherence of aspirin users

```{r}
cdm <- generateIngredientCohortSet(
  cdm = cdm,
  name = "aspirin",
  ingredient = "aspirin",
  gapEra = 7
)

survivalResultAspirin <- estimateSingleEventSurvival(
  cdm,
  "aspirin",
  "aspirin",
  outcomeDateVariable = "cohort_end_date"
)

survivalResultAll <- omopgenerics::bind(
  survivalResult,
  survivalResultAspirin
)
```

## Exercise 1 - Estimate treatment adherence of aspirin users

```{r}
riskTable(survivalResultAll)
```

## Exercise 1 - Estimate treatment adherence of aspirin users

```{r}
plotSurvival(survivalResultAll, colour = "outcome")
```

# Stratification

## Adding strata - Example

::: {style="margin-bottom: 10px;"}
:::

-   We can stratify our survival study by any variables available in the target cohort provided. We can use `PatientProfiles` to add, for instance, demographic information.

. . .

```{r}
cdm$ibuprofen_strata <- cdm$ibuprofen |> 
  PatientProfiles::addAge(
    ageGroup = list(c(0,40),c(41,70),c(71,150))
  ) |>
  compute(name = "ibuprofen_strata")

cdm$ibuprofen_strata |>
  dplyr::select(-"age") |>
  head()
```

## Adding strata - Example

```{r}
survivalResultStrata <- estimateSingleEventSurvival(
   cdm,
  "ibuprofen_strata",
  "ibuprofen_strata",
  outcomeDateVariable = "cohort_end_date",
  strata = list("age_group")
)
```

## Adding strata - Example

```{r, fig.height = 9}
plotSurvival(survivalResultStrata, colour = "age_group", riskTable = TRUE)
```

# Your turn!

## Exercise 2 - Stratification

**Use the same aspirin cohort as before.**

-   Add both sex and age_group information.

-   Estimate treatment adherence and plot all Kaplan-Meier curves in one plot.

## Exercise 2 - Stratification

```{r}
cdm$aspirin_strata <- cdm$aspirin |> 
  PatientProfiles::addDemographics(
    ageGroup = list("kids" = c(0,18), "adults" = c(19,150))
  ) |>
  compute(name = "aspirin_strata")

survivalResultStrata <- estimateSingleEventSurvival(
   cdm,
  "aspirin_strata",
  "aspirin_strata",
  outcomeDateVariable = "cohort_end_date",
  strata = list("age_group", "sex", c("age_group", "sex"))
)
```

## Exercise 2 - Stratification

```{r}
plotSurvival(survivalResultStrata, colour = "age_group", facet = "sex")
```

# Additional input choices

## Let's play with the parameters of the survival estimation function

Reminder of all possible ways we can tweak our analysis:

```{r, eval = FALSE}
# don't run
estimateSingleEventSurvival(
  cdm = cdm,
  targetCohortTable = "ibuprofen",
  outcomeCohortTable = "ibuprofen",
  outcomeDateVariable = "cohort_end_date",
  outcomeWashout = Inf,
  censorOnCohortExit = FALSE,
  censorOnDate = NULL,
  followUpDays = Inf,
  strata = NULL,
  eventGap = 30,
  estimateGap = 1,
  restrictedMeanFollowUp = NULL,
  minimumSurvivalDays = 1
)
```

## Changing inputs - Example

::: {style="margin-bottom: 10px;"}
:::

Let's change the event gap to a weekly aggregation and display the risk table under the plot.

. . .

```{r}
survivalResultEventGap <- estimateSingleEventSurvival(
   cdm,
  "ibuprofen",
  "ibuprofen",
  outcomeDateVariable = "cohort_end_date",
  eventGap = 7
)
```

## Changing inputs - Example

```{r, fig.height = 6}
plotSurvival(survivalResultEventGap, riskTable = TRUE, riskInterval = 7)
```

# Your turn!

## Exercise 3 - Changing inputs

**Keep working with the cohort of individuals starting aspirin prescription.**

-   Now estimate survival with a 5-year washout period, instead of considering first events.

## Exercise 3 - Changing inputs

```{r}
survivalResultAspirinTuned <- estimateSingleEventSurvival(
   cdm,
  "aspirin",
  "aspirin",
  outcomeDateVariable = "cohort_end_date",
  outcomeWashout = 1825
)

omopgenerics::bind(survivalResultAspirin, survivalResultAspirinTuned) |> 
  tableSurvival()
```

## Exercise 3 - Changing inputs

```{r}
plotSurvival(omopgenerics::bind(survivalResultAspirin, survivalResultAspirinTuned), 
             colour = "outcome_washout")
```

## Exercise 4 - Survival with different exposure and outcome

What is the time to first myocardial infraction after starting simvastatin or aspirin treatment, and how does this differ between both prescription drugs?

## Exercise 4 - MI after start of simvastatin or aspirin

```{r}
conceptSetMI <- CodelistGenerator::getCandidateCodes(cdm, keywords = "myocardial")
mi <- list("mi" = conceptSetMI$concept_id)
cdm$mi <- CohortConstructor::conceptCohort(
  cdm, conceptSet = mi, name = "mi"
)

cdm <- DrugUtilisation::generateIngredientCohortSet(
  cdm, 
  name = "statins",
  ingredient = c("simvastatin", "aspirin"),
  gapEra = 7
)

survivalStatin <- estimateSingleEventSurvival(
  cdm,
  "statins",
  "mi"
)
```

## Exercise 4 - MI after start of simvastatin or aspirin

```{r}
plotSurvival(survivalStatin, colour = "target_cohort")
```

# Thank you!

**Questions?**
